---
title: "Dynamics Foundations of the Valence Foundry"
author: "Ed Phillips, Jan Ritch-Frel, Flow Feel"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dynamics Foundations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(valence.foundry)
library(ggplot2)
```

## Overview

This document gives the mathematical foundations of the **dynamics module** of
the valence (Valence-Ingression) framework. Three families of dynamics are treated:

1. The **cusp catastrophe** — how a bistable equilibrium `x³ + ax + b = 0`
   produces hysteresis and irreversibility.
2. **Autocatalytic growth** — why the per-capita innovation rate *rises* with
   diversity (the "Homo inversion") and why growth stays bounded.
3. **Cross-kingdom transfer** — what transfers between biological and cultural
   domains (the *sign* of a pattern) and what does not (its *magnitude*).
4. **Economics** — commitment-driven (CDI) option-value decay and stochastic
   CDI paths.

Every result that is *proved* here is returned as a `valence_proof` object
(statement, LaTeX-like derivation, `QED` result, a `verified` flag and a
`numeric_check` cab), and every module result is wrapped in an S3 class
(`valence_cusp_result`, `valence_autocatalytic_result`, `valence_economics_result`) built on
the shared `valence_dynamics_result` base — the Phosphene R Standards §7
three-layer pattern (constructor → validator → helper).

All functions obey the DFT axioms: **A1** (pure IO separation), **A2**
(determinism), **A6** (check-result with provenance).

---

## 1. The Cusp Catastrophe

### 1.1 The equilibrium

The dynamics module tracks a system whose equilibrium solves the canonical
cusp normal form

$$x^3 + a\,x + b = 0,$$

where $x$ is the state (e.g. an attained diversity level or a commitment) and
$b$ is the swept control parameter, while $a$ sets the shape of the response
surface.

```{r cusp-first-look}
proof_pos <- prove_cusp_bifurcation(a = 1, b = 0.3)
proof_pos$verified               # no bifurcation for a > 0
proof_neg <- prove_cusp_bifurcation(a = -1, b = 0.2)
proof_neg$verified               # bistable (3 real roots) for a < 0
```

### 1.2 The bifurcation set

The discriminant of the cubic $x^3 + p x + q$ is
$\Delta = -4p^3 - 27q^2$. With $p = a$ and $q = b$:

$$\Delta(a,b) = -4a^3 - 27b^2.$$

* **Case a > 0.** Since $-4a^3 < 0$ and $-27b^2 \le 0$, always $\Delta < 0$:
  exactly **one real root**, so the response is single-valued and there is no
  bifurcation.
* **Case a < 0.** Now $-4a^3 > 0$, so $\Delta > 0$ precisely when
  $27b^2 < -4a^3$, i.e.

$$|b| < b_{\mathrm{crit}} := \frac{2}{3\sqrt{3}}\,(-a)^{3/2},$$

which yields **three real roots** — the bistable regime. The boundary where
the cubic acquires a double root is the **bifurcation set**:

$$\boxed{\,4a^3 + 27b^2 = 0\,}$$

```{r cusp-bif-set}
b_crit <- (2 / (3 * sqrt(3))) * 1^(3 / 2)   # for a = -1
b_crit
```

### 1.3 Hysteresis derivation

Within the bistable region there are two *stable* branches (an upper and a
lower) separated by an unstable middle branch. When the control $b$ is ramped
up, branch-following keeps the state on the upper branch until it collapses at
the upper saddle-node; when $b$ is ramped back down, the state persists on the
lower branch. The forward and reverse sweeps therefore trace **different**
paths — path dependence, i.e. **hysteresis**.

The hysteresis loop *area* is the integral of the branch separation over the
bistable range,

$$A = \int_{-b_{\mathrm{crit}}}^{b_{\mathrm{crit}}}
   \big(x_{\mathrm{hi}}(b) - x_{\mathrm{lo}}(b)\big)\,db,$$

where $x_{\mathrm{hi}}$ and $x_{\mathrm{lo}}$ are the two outer real roots of
the cubic. This analytical area is reproduced by the trapezoid-rule loop area
recovered from the numeric forward/reverse sweeps.

```{r cusp-loop}
loop_proof <- prove_hysteresis_loop_area(a = -1, control_range = c(-2, 2), n = 200)
loop_proof$verified   # numeric area matches analytic area
loop_proof$numeric_check
```

```{r cusp-plot, fig.width=6, fig.height=4}
eq_fn <- make_cusp_equilibrium_fn(a = -1)
cv <- seq(-2, 2, length.out = 200)
res <- cusp_hysteresis_check(cv, eq_fn, seed = 42L)
df <- data.frame(
  control  = rep(cv, 2),
  state    = c(res$values$forward_states, res$values$reverse_states),
  sweep    = rep(c("forward", "reverse"), each = length(cv))
)
ggplot(df, aes(control, state, colour = sweep)) +
  geom_line() +
  labs(title = "Hysteresis loop of the cusp (a = -1)",
       x = "control b", y = "equilibrium state x") +
  theme_minimal()
```

---

## 2. Autocatalytic Growth

### 2.1 Per-capita rate increases with diversity

Consider autocatalytic growth

$$\frac{dN}{dt} = r\,N\left(\frac12 + \frac12\,\frac{N}{N+K}\right),$$

so the per-capita rate is

$$f(N) = r\left(\frac12 + \frac12\,\frac{N}{N+K}\right).$$

Differentiating,

$$\frac{df}{dN} = \frac{r}{2}\,\frac{d}{dN}\!\left(\frac{N}{N+K}\right)
= \frac{r}{2}\,\frac{K}{(N+K)^2} > 0 \qquad (N>0).$$

The per-capita rate is **monotonically increasing** — the "Homo inversion".
By contrast the logistic per-capita rate $g(N)=r(1-N/K)$ has
$dg/dN = -r/K < 0$: niche-filling predicts a *decreasing* per-capita rate.
This sign flip is exactly what distinguishes valence from its competitor.

```{r dd-sign}
dd_proof <- prove_dd_sign(r = 1, K = 10, N_max = 100, n_grid = 100)
dd_proof$verified
dd_proof$numeric_check   # min d/dN of per-capita rate (positive)
```

### 2.2 Boundedness

At $N=0$ the rate is $f(0) = r/2$; as $N \to \infty$, $N/(N+K)\to 1$ so
$f(N) \to r$. Since $0 \le N/(N+K) < 1$, we always have
$r/2 \le f(N) < r$:

* the rate is **bounded below** by $r/2$ and **above** by $r$ (no blow-up);
* growth is at most linear in $N$, so counts cannot explode in finite time.

```{r bounded}
bnd_proof <- prove_autocatalytic_growth_rate(r = 1, K = 10, N_max = 5000, n_grid = 100)
bnd_proof$verified
```

```{r growth-plot, fig.width=6, fig.height=4}
NN <- seq(0, 200, length.out = 400)
df2 <- data.frame(
  N = NN,
  rate = 1 * (0.5 + 0.5 * NN / (NN + 10))
)
ggplot(df2, aes(N, rate)) +
  geom_hline(yintercept = c(0.5, 1), lty = 2, colour = "grey40") +
  geom_line(colour = "#C0392B", linewidth = 1) +
  labs(title = "Autocatalytic per-capita rate: 0.5r up to r (bounded)",
       y = "per-capita rate f(N)") +
  theme_minimal()
```

The `valence_autocatalytic_result` class packages these quantities. We can coerce a
raw `diversity_dependence_sign()` output into the classed result:

```{r auto-class}
counts <- c(1, 2, 4, 8, 15, 28, 52, 96, 180, 340)
raw <- diversity_dependence_sign(counts, seed = 42L)
res <- as_valence_autocatalytic_result(raw)
res$values$diversity_dependence_sign
summary(res)
```

---

## 3. Cross-Kingdom Transfer

The inference framework is *scaled across kingdoms*: a pattern that is an
inversion-invariant **sign** (e.g. "positive diversity-dependence", "negative
option decay in commitment") is expected to reproduce across biological and
cultural substratescars — **the sign transfers**. What does **not** transfer is
**magnitude**: the numeric slope or area depends on domain-specific units,
time scales, and carrying capacities.

Viewed through the cusp, the *shape* of the bifurcation (presence/absence of
hysteresis, positive loop area) is the transferable invariant; the specific
$a,K,r$ are per-system. Proof functions verify a **sign/structure** invariant,
never a magnitude, which is why `prove_dd_sign()` proves a *sign* (positive vs
negative) and `prove_autocatalytic_growth_rate()` proves *boundedness* (a
qualitative, magnitude-free property).

The cross-kingdom expectation is therefore:

| Property | Transfers? | Why |
|----------|-----------|-----|
| Sign of diversity-dependence | Yes | sign is inversion-invariant |
| Presence of hysteresis | Yes | structural, magnitude-free |
| Boundedness of growth | Yes | qualitative, no units |
| Numeric per-capita slope | No | domain-specific magnitude |
| Loop area (absolute) | No | units depend on control scaling |

---

## 4. Economics

### 4.1 CDI trajectory

Commitment-driven disruptions are described by a *commitment-diversity index*
(CDI) trajectory that accrues fast early and decelerates as it saturates —
commitment is a bounded sum of committed capacity, so CDI lives in
$[0,1]$.

### 4.2 Option-value decay

Standard options theory decays the value of the pivot option exponentially in
*time*; valence couples decay to *commitment*: because remaining capacity is
$1 - \text{CDI}$, the option value falls monotonically as CDI rises.

```{r econ-opt}
cdi_grid <- seq(0, 1, length.out = 30)
k <- 2
option_value <- exp(-k * cdi_grid)
econ <- valence_economics_result(
  cdi = cdi_grid,
  option_value = option_value,
  stochastic_paths = matrix(rep(cdi_grid, 3), ncol = 3),
  threshold_disruption = cdi_grid > 0.8
)
all(diff(econ$values$option_value) <= 0)   # non-increasing option value
```

### 4.3 Stochastic CDI paths

Because CDI is a bounded quantity it is sensible to model it as a
mean-reverting, *bounded* stochastic process: each simulated path is clamped
to $[0,1]$.

```{r econ-stoch}
set.seed(42)
sres <- stochastic_cdi(mu0 = 0.5, sigma0 = 0.2, cdi_init = 0.01,
                       n_steps = 200L, threshold = 0.8, seed = 42L)
path <- sres$metadata$path
range(path)           # within [0, 1]
```

```{r econ-plot, fig.width=6, fig.height=4}
df3 <- data.frame(
  step = seq_along(path),
  cdi  = path
)
ggplot(df3, aes(step, cdi)) +
  geom_hline(yintercept = c(0, 1), lty = 2, colour = "grey40") +
  geom_line(colour = "#2E86C1") +
  labs(title = "Stochastic CDI path (bounded in [0, 1])",
       x = "time step", y = "CDI") +
  ylim(0, 1) +
  theme_minimal()
```

The `valence_economics_result` class bundles the CDI trajectory, option-value
schedule, stochastic paths, and disruption outcome, and — like all
dynamics results — prints, summarises to a tidy data.frame, and converts via
`as.data.frame()`.

---

## Summary

The dynamics module is mathematically closed at three levels:

* **Cusp catastrophe** — bifurcation set $4a^3 + 27b^2 = 0$, hysteresis
  derived from branch-following, loop area verified analytically vs
  numerically.
* **Autocatalytic growth** — strictly increasing per-capita rate (positive
  DD), bounded between $r/2$ and $r$: the substrate for the Homo inversion.
* **Economics** — bounded CDI, monotone option decay in commitment.

Because the proofs fix *signs and structure* rather than magnitudes, their
conclusions transfer across kingdoms while the specifics are per-system — the
central epistemology of the valence framework.