---
title: "Exploring the framework in Toy Models"
author: "Ed Phillips"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring the framework in Toy Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
library(valence.foundry)
```

## Scientific motivation

The empirical tests in the the foundry are blocked on data (Review Items 4–6:
gene-category data, island-bird morphology, hominin/economic time series).
The formal model is a theoretical ODE that cannot fail — it simulates the framework's
equation and produces the framework's prediction by construction. Between these lies a
gap: a layer that lets a reader *explore* the consequences of the framework across
parameter space and hypothetical substrates, without claiming empirical test.

The **toy models** fill this gap. Each realm is a self-contained "what if"
world with a simulator, a set of dials, and a visualization. They ask: *if the framework
were true, what would we expect to see in worlds we have not measured?* The
answer sharpens the predictions and, when the data arrives, makes the
empirical tests more discriminating.

This is **not** empirical testing. The realms do not source new data, do not
claim to corroborate the framework, and do not replace the blocked empirical work. Each
realm names the experiment that would convert it from speculative to
empirical. See `docs/review/toy-models-plan.md` for the full execution plan.

## Method overview

| Realm | What it explores | Reuses | Status |
|-------|-----------------|--------|--------|
| 1. Genome reduction | The threshold gate as a control parameter | `threshold_model()` | Built |
| 2. Irreversibility | Hysteresis loop area vs. cusp parameter `a` | `cusp_hysteresis_check()` | Planned |
| 3. *Homo* inversion | Diversity-dependence sign flip | `generate_autocatalytic_set()` | Planned |
| 4. Cross-kingdom | When model transfer beats sign-only transfer | `empirical_formal_model()` | Planned |

## Realm 1 — The threshold gate

### What it explores

The threshold gate is the heart of the framework's biphasic prediction. The formal model
([`threshold_model()`](../reference/threshold_model.html)) integrates:

$$\frac{dC_i}{dt} = -\lambda \cdot M(t) \cdot C_i \cdot \mathbb{1}(d_i < \theta)$$

where $C_i$ is the retention probability of trait $i$, $M(t) = M_0 e^{-\alpha t}$
is the decaying niche-demand mismatch, $d_i$ is the trait's integration depth,
and $\theta$ is the protection threshold. Traits with $d_i \geq \theta$ are
protected (retention = 1.0); traits with $d_i < \theta$ are shed
(retention $\to 0$).

The **biphasic signal** is the gate itself: protected traits retain at 1.0
while unprotected traits shed to ~0, producing a sharp contrast
(`threshold_biphasicity` ≈ 1.0). This is the real biphasic signature — not
the temporal displacement ratio (math-review Issue 3, resolved).

This realm makes the gate *visible* as a control parameter. By sweeping
$\theta$ across a grid, we can watch the gate open (biphasicity → 1) when
$\theta$ separates protected from unprotected traits, and close (biphasicity
→ 0) when $\theta$ is below the minimum depth (all protected) or above the
maximum depth (all unprotected).

### Worked example

Consider a hypothetical organism with five traits at integration depths
`r paste(c(0, 1, 2, 3, 5), collapse = ", ")`. We sweep $\theta$ from 0 to 6:

```{r sweep}
result <- sweep_threshold(
  depths = c(0, 1, 2, 3, 5),
  theta_grid = seq(0, 6, by = 0.25)
)
```

The sweep returns a data frame with one row per $\theta$ value:

```{r sweep-head}
head(result$values$sweep, 8)
```

The peak biphasicity is `r result$values$peak_biphasicity` at
$\theta = `r result$values$peak_theta`$. This is where the gate is maximally
open — the cleanest separation between protected and unprotected traits.

```{r plot-gate, fig.alt = "Threshold gate: biphasicity vs theta. The gate opens at theta > 0, stays at 1.0 through the interior, and closes at theta > 5."}
plot_threshold_gate(result)
```

The shaded region is the "gate open" zone — the parameter range where the framework's
biphasic prediction is active. The vertical dashed lines mark each trait's
depth; the gate composition changes at each one.

### What this teaches

1. **The biphasic signal IS the gate.** The `threshold_biphasicity` metric
   captures the protected-vs-unprotected contrast. It is 1.0 whenever both
   classes are non-empty, and 0 when one class is empty. This is the signal
   the math-review (Issue 3) identified as the real biphasic signature — not
   the temporal displacement ratio.

2. **The gate is a step function.** The biphasicity rises sharply from 0 to 1
   as $\theta$ crosses the minimum depth, stays at 1 through the interior,
   and falls sharply to 0 as $\theta$ crosses the maximum depth. The gate is
   binary — it is either open or closed.

3. **The R6 method-misspecification is visceral.** A cross-sectional
   regression across organisms with *different* $\theta$ values would see
   noise, not a logistic — which is exactly why T3 fails on real
   endosymbiont data. Different lineages have different thresholds (different
   dependency architectures, different parasitism levels), so pooling them
   cross-sectionally destroys the within-lineage biphasic signal.

### Skewing the dependency architecture

The gate's shape depends on the dependency architecture. A uniform
architecture (`c(0, 1, 2, 3, 5)`) opens the gate gradually. A skewed
architecture — many shallow traits, few deep — changes the composition:

```{r skewed}
skewed <- sweep_threshold(
  depths = c(0, 0, 0, 1, 5, 5),
  theta_grid = seq(0, 6, by = 0.25)
)
plot_threshold_gate(skewed)
```

The gate still opens and closes, but the protected/unprotected counts change
differently. At $\theta = 0.5$, three traits are unprotected (the three at
depth 0) and three are protected — a different composition from the uniform
case.

### The experiment that would fill this realm

A real organism with a known dependency architecture (gene categories +
integration-depth scores) and a measured retention trajectory across a
parasitism/commitment gradient. The $\theta$-sweep would then be calibrated
against real data, and the gate's $\theta$ would be an empirical estimate,
not a dial. The foundry's `empirical_formal_model()` (Remark R7) is a step
in this direction — it fits the retention matrix to estimate $dep$ and $para$
effects — but the full gate calibration requires gene-category data (Item 4).

## Realm 2 — Irreversibility

### What it explores

Irreversibility is the framework's sharpest departure from gradual reversibility. The
[cusp catastrophe](../reference/cusp_hysteresis_check.html) models a system
where smooth changes in a control parameter produce sudden, irreversible
jumps in state. The cusp equilibrium is:

$$x^3 + a \cdot x + b = 0$$

where $x$ is the state, $a$ is the splitting factor, and $b$ is the normal
factor. For $a < 0$, the system has two stable equilibria (a cusp); crossing
the bifurcation set ($4a^3 + 27b^2 = 0$) forces a jump. For $a \geq 0$, there
is a single equilibrium (no bifurcation, fully reversible).

The **hysteresis loop area** quantifies irreversibility: if you sweep $b$
forward (increasing) then reverse (decreasing), the forward and reverse paths
enclose a loop. The area of that loop is zero for a reversible system ($a \geq
0$) and positive for an irreversible system ($a < 0$), growing as $|a|$
increases.

### Worked example

We sweep the cusp parameter $a$ from 1 (no bifurcation) through 0 (the
bifurcation point) to $-2$ (deep in the cusp region), computing the loop
area at each value:

```{r sweep-irrev}
irrev <- sweep_cusp_irreversibility(
  a_grid = seq(1, -2, by = -0.1),
  control_values = seq(-2, 2, length.out = 100)
)
```

The sweep returns a data frame with one row per $a$ value:

```{r irrev-head}
head(irrev$values$sweep, 6)
tail(irrev$values$sweep, 6)
```

The peak loop area is `r round(irrev$values$peak_loop_area, 3)` at $a =
`r irrev$values$peak_a` — the most negative $a$ in the grid, where the cusp
region is widest.

```{r plot-irrev, fig.alt = "Irreversibility sweep: loop area vs a. The area is 0 for a >= 0 and rises for a < 0, showing irreversibility emerging at the bifurcation."}
plot_irreversibility_sweep(irrev)
```

The shaded region is the cusp region ($a < 0$) — where the framework's irreversibility
prediction is active. The dashed line marks the bifurcation ($a = 0$). For
$a \geq 0$, the loop area is exactly 0 (the system is fully reversible). For
$a < 0$, the loop area rises as $|a|$ grows — the cusp region widens, and the
hysteresis loop grows.

### What this teaches

1. **Irreversibility is quantitative, not boolean.** `cusp_hysteresis_check()`
   returns a boolean (`has_hysteresis`), but the loop area is a continuous
   measure of *how much* irreversibility there is. A system at $a = -0.5$ has
   a small loop; a system at $a = -2$ has a large one.

2. **Irreversibility emerges at the bifurcation.** The loop area is exactly 0
   for $a \geq 0$ and rises discontinuously for $a < 0$. This is the
   bifurcation — the point where the framework predicts irreversibility emerges.

3. **The loop area is robust to the path.** The loop area is the same
   regardless of which branch you start on (the loop encloses the same area
   either way). This makes the metric reliable: it depends only on the system
   ($a$), not on the observer's starting point.

### The experiment that would fill this realm

A real system (a lineage, an ecosystem, an economy) where the control
parameter (parasitism depth, commitment, complexity) has been ramped both up
and down, with the state (genome size, capacity, diversity) measured at each
point. The loop area would then be an empirical quantity. No such experiment
exists for capacity reallocation; the realm makes this absence visible — the
reader can see exactly what data would count as the test.

## Realm 3 — The Homo inversion

### What it explores

The *Homo* inversion is the framework's signature theoretical contribution:
the prediction that, after a substrate shift, innovations generate further
innovations faster than they are lost — *positive diversity-dependence* (the
per-capita innovation rate increases with diversity $N$). This is distinct
from standard niche-filling, where the per-capita rate *decreases* with $N$
(negative diversity-dependence, logistic growth).

This realm makes the DD sign flip *visible*. It generates two matched
diversity time series — one autocatalytic (positive DD, the *Homo* inversion)
and one logistic (negative DD, niche-filling) — and overlays them. Both
trajectories grow (positive growth direction), but their per-capita-rate-vs-$N$
slopes have opposite signs: that is the *Homo* inversion.

It also explores the **endogenous-K bifurcation** (Review Item 3): the
prediction that there is a threshold in cultural complexity beyond which DD
flips from negative (niche-filling) to positive (autocatalytic). Sweeping the
feedback parameter shows this flip.

### Risk note: the corrected feedback formula

The execution plan (`docs/review/toy-models-plan.md`) proposed the per-capita
rate formula $r \times (\text{feedback} + (1-\text{feedback}) \times N/(N+K))$,
intended to interpolate between logistic (feedback = 0, negative DD) and
autocatalytic (feedback = 0.5, positive DD). **This formula is wrong**: at
feedback = 0 it gives $r \times N/(N+K)$, whose derivative w.r.t. $N$ is
$rK/(N+K)^2 > 0$ — always positive DD. The function $N/(N+K)$ is *increasing*
in $N$, so the blend is always increasing; it cannot produce negative DD at
any feedback value.

The workaround blends two bounded functions — one increasing (autocatalytic),
one decreasing (bounded logistic):

$$\text{pc\_rate} = r \times [\text{feedback} \times (0.5 + 0.5 N/(N+K)) + (1-\text{feedback}) \times K/(N+K)]$$

The analytical derivative is $rK/(N+K)^2 \times (1.5\,\text{feedback} - 1)$:
negative for feedback $< 2/3$ (niche-filling), positive for feedback $> 2/3$
(*Homo* inversion), zero at feedback $= 2/3$ (the bifurcation). At feedback = 1
this reduces to $r(0.5 + 0.5N/(N+K))$, exactly matching the existing
`generate_autocatalytic_set()` dynamics (verified: identical time series).

### Worked example: the contrast

```{r dd-contrast}
contrast <- diversity_dependence_contrast(n_steps = 20, capacity = 30)
```

The autocatalytic DD slope is `r round(contrast$values$autocatalytic_dd_slope, 6)`
(positive — the *Homo* inversion); the logistic DD slope is
`r round(contrast$values$logistic_dd_slope, 6)` (negative — niche-filling). The
contrast is `r round(contrast$values$contrast, 6)` — positive, confirming the
The framework signature.

```{r plot-dd-contrast, fig.alt = "Two-panel plot: left shows autocatalytic and logistic trajectories both growing; right shows per-capita rate vs N with opposite slopes (autocatalytic increasing, logistic decreasing)."}
plot_dd_contrast(contrast)
```

The left panel shows both trajectories growing — they look similar. The right
panel reveals the difference: the autocatalytic per-capita rate *increases*
with $N$ (green, positive slope), while the logistic per-capita rate
*decreases* with $N$ (red, negative slope). That slope flip is the *Homo*
inversion.

### Worked example: the endogenous-K sweep

```{r dd-sweep}
dd_sweep <- sweep_endogenous_k(feedback_grid = seq(0, 1, by = 0.05))
```

The sweep shows the DD sign flipping from negative to positive:

```{r dd-sweep-table}
head(dd_sweep$values$sweep, 8)
tail(dd_sweep$values$sweep, 8)
```

The theoretical bifurcation is at feedback = `r round(dd_sweep$values$bifurcation_feedback, 4)`
(2/3); the measured bifurcation (where the linear regression slope crosses
zero) is `r round(dd_sweep$values$measured_bifurcation_feedback, 4)` — very
close to the theoretical value.

### What this teaches

1. **The DD sign is the *Homo* inversion signature.** Both logistic and
   autocatalytic growth produce growing time series — growth direction alone
   cannot distinguish them. The per-capita-rate-vs-$N$ slope is the
   discriminator: positive = *Homo* inversion, negative = niche-filling.

2. **The bifurcation is quantitative.** The DD slope doesn't just flip; it
   crosses zero continuously. The endogenous-K bifurcation is a threshold in
   cultural complexity (the feedback parameter), and the sweep shows exactly
   where it is (feedback $\approx 2/3$).

3. **The measured bifurcation is approximate.** The linear regression in
   `diversity_dependence_sign()` fits a linear approximation to a nonlinear
   per-capita-rate function. The sign is reliable, but the exact crossing
   point may differ slightly from the theoretical $2/3$. This is documented as
   a known limitation.

### The experiment that would fill this realm

A hominin (or any cultural-species) diversification time series spanning the
putative complexity threshold: innovation counts or cultural-variant counts
through time, long enough to estimate the per-capita-rate-vs-diversity slope
on both sides of the threshold. No such dataset exists; the realm makes the
prediction and its data requirement concrete.

## Realm 4 — The cross-kingdom transfer

### What it explores

The cross-kingdom transfer is the monograph's strongest claim: a model fit to
one kingdom (plants) predicts the ordering in another (birds). But the
foundry's `transfer_test()` transfers only the slope *sign* — ranking
discards magnitude by construction (math-review Issue 7). And the empirical
GLM (`empirical_formal_model()`) forces all birds to the same parasitism
level, which makes *its* transfer sign-only too: when para is constant,
$\text{rank}(\text{logistic}(a + b \cdot \text{dep} + c \cdot \text{const})) =
\text{rank}(\text{dep})$ if $b > 0$.

This realm asks: **when would the full model (both coefficients + the
logistic link) transfer genuinely more than the sign alone?** Answer: when
the target kingdom has *varying* parasitism scores, so the model can use its
para coefficient to modulate predictions — information the sign-only transfer
(which ignores para) cannot access.

### Risk note: why this is speculative

The synthetic bird data has *varying* parasitism scores — this is what makes
the model transfer differ from the sign-only transfer. The **real** bird data
(`island_bird_morphology.csv`) has no parasitism column, so
`empirical_formal_model()` forces all birds to the same para level, making its
transfer sign-only by construction. This realm is therefore speculative: it
shows what the transfer *could* look like with richer target-kingdom data, not
what it does look like with the data we have.

### Worked example: model vs sign-only at low noise

```{r glm-transfer}
d <- valence.foundry:::generate_transfer_data(noise_sd = 0.05, seed = 42L)
transfer <- glm_transfer(d$plant_data, d$bird_data, seed = 42L)
```

The model-transfer rho is `r round(transfer$values$model_rho, 4)`; the
sign-only-transfer rho is `r round(transfer$values$sign_only_rho, 4)`. The
model advantage is `r round(transfer$values$model_advantage, 4)` — positive,
confirming the model outperforms the sign at low noise. The GLM recovers the
dep coefficient as `r round(transfer$values$dep_coefficient, 4)` (true: 0.8)
and the para coefficient as `r round(transfer$values$para_coefficient, 4)`
(true: -0.5).

### Worked example: the noise sweep

```{r transfer-sweep}
transfer_sweep <- sweep_transfer_robustness(
  noise_grid = seq(0, 0.3, by = 0.05)
)
```

```{r transfer-sweep-table}
transfer_sweep$values$sweep
```

The sign-only rho is flat at `r round(transfer_sweep$values$sweep$sign_only_rho[1], 4)`
across all noise levels — it depends only on the dep sign, which doesn't flip.
The model rho starts at `r round(transfer_sweep$values$sweep$model_rho[1], 4)`
(zero noise) and degrades to `r round(tail(transfer_sweep$values$sweep$model_rho, 1), 4)`
(at noise = 0.3), converging to the sign-only rho. The low-noise advantage is
`r round(transfer_sweep$values$low_noise_advantage, 4)`; convergence occurs at
noise = `r round(transfer_sweep$values$convergence_noise, 4)`.

```{r plot-transfer-sweep, fig.alt = "Plot of noise vs rho, showing the model-transfer curve starting high and degrading down to the flat sign-only-transfer line."}
plot_transfer_breakdown(transfer_sweep)
```

The gap between the two curves is the *information lost to ranking* (Issue 7):
the magnitude the sign-only transfer discards. As noise increases, the gap
closes — the model degrades to the sign.

### What this teaches

1. **Ranking discards magnitude (Issue 7), visible as a gap.** The sign-only
   transfer is flat across noise (it only needs the sign); the model transfer
   degrades (its coefficient estimates get noisier). The gap is the magnitude
   information the sign-only transfer throws away.

2. **The model transfer is not unconditionally superior.** At extreme noise
   (beyond the default grid, e.g. 0.4–0.5), the model's noisy para
   coefficient injects noise that the sign-only transfer avoids — the model
   can dip *below* the sign-only transfer. The model pays a noise penalty for
   using the para coefficient. This is an honest finding: the model transfer
   outperforms only when the GLM coefficients are well-estimated.

3. **The real bird data makes the empirical transfer sign-only.** With no
   parasitism column in the real bird data, `empirical_formal_model()` forces
   all birds to the same para level — making the empirical transfer
   sign-only by construction. Richer target-kingdom data (with varying
   parasitism) would let the model transfer genuinely outperform.

### The experiment that would fill this realm

A real target-kingdom dataset with *varying* parasitism/commitment scores
(not just dependency scores): e.g., bird morphology data with a per-species
parasitism or commitment-level column, or a third substrate (economic
specialization) with varying commitment. The GLM transfer would then be
empirical, and the model-vs-sign gap would be a real, testable quantity —
not a synthetic one. The foundry now has the plant data (R7) and the bird
data (bundled); the varying-parasitism column is the open empirical frontier.

## Limitations

The toy models are **theoretical exploration, not empirical test**:

- They do not source new data.
- They do not claim to corroborate the framework.
- They do not replace the blocked empirical work (Items 4–6).
- The sweep uses the default ODE parameters ($\lambda = 0.15$, $M_0 = 10$,
  $\alpha = 0.05$, $t = 100$); with different parameters, the unprotected
  traits may not fully shed, and the biphasicity would be less than 1.0 in
  the interior. The gate shape (0 → contrast → 0) is robust, but the
  plateau height is parameter-dependent.

## Interpretation guide

| What you see | What it means |
|-------------|---------------|
| Biphasicity = 0 at $\theta = 0$ | All traits protected — no contrast (gate closed) |
| Biphasicity = 1 in the interior | Gate open — clean separation between protected and unprotected |
| Biphasicity = 0 at $\theta > \max(d)$ | All traits unprotected — no contrast (gate closed) |
| Vertical dashed lines | Trait depths — gate composition changes here |
| Peak $\theta$ | The threshold that maximally separates the two classes |

The key insight: **the threshold gate is the biphasic signal**. When the
gate is open (both classes exist), the framework's biphasic prediction is active. When
the gate is closed (one class empty), there is no biphasic contrast. This is
why T3's cross-sectional regression fails — it pools organisms with different
gates, averaging the contrast away.
