---
title: "The Foundry: Testing the framework Predictions with R"
author: "Ed Phil, Jan Ritch-Frel, Flow"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Foundry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Scientific Motivation

## The framework under review

the framework is a formal theory of capacity reallocation under
niche-driven selection. The core claim is that when an organism commits to a
novel niche, the rate at which it sheds ancestral functions follows a
**threshold-gated integration-depth ordering**:

1. **Integration depth** ($d_i$) — how deeply a trait is wired into the
   organism's regulatory and functional architecture — determines its
   resistance to loss.
2. Traits with $d_i < \theta$ (the protection threshold) are shed rapidly
   during Phase 1 of niche commitment.
3. Traits with $d_i \ge \theta$ are retained indefinitely, even under
   extreme niche specialisation.
4. The ordering of trait loss across the genome (or phenome) is predictable
   from integration depth: deeply integrated functions are lost last.

This is formalised as the dynamical model:

$$\frac{dC_i}{dt} = -\lambda \cdot M(t) \cdot I(d_i < \theta)$$

where $C_i$ is the retention probability of trait $i$, $M(t)$ is an
exponentially decaying niche-demand mismatch, $\lambda$ is the shedding
rate, and $\theta$ is the protection threshold.

## What the framework predicts

| Test | Prediction | In Brief |
|------|-----------|----------|
| **T1** | Plastome size ~ parasitism level | Orobanchaceae PGLS |
| **T2** | Cross-family gradient replication | 12+ parasitic plant families |
| **T3** | Biphasic genome reduction | Endosymbiont logistic decay |
| **T4** | Niche breadth > Ne for gene loss | Bobay & Ochman (2017) |
| **T5** | Pan-genome fluidity ~ lifestyle | Dewar et al. (2024) |
| **T6** | Integration-depth determines loss order | Exact permutation test |
| **T7** | Function-loss co-segregation with sweeps | LTEE (Good et al. 2017) |
| **L3** | Cross-kingdom parameter transfer | Plants → birds |

## The Gap

Existing theories make overlapping predictions:

- **Relaxed selection** (Lahti et al. 2009) predicts the same loss gradient
  as the framework for T1–T2, but through a different mechanism.
- **Genetic drift** (Lynch 2007) predicts $\mathrm{N_e}$ is the primary
  driver of genome reduction — contradicting T4–T5.
- **Muller's ratchet** predicts accelerating (not decelerating) loss —
  contradicting T3.
- **McCutcheon's metabolic complementarity** predicts the same correlation
  as T3 but does not predict cross-kingdom transfer (L3).
- **Random loss** predicts no integration-depth ordering — contradicted by
  T6.
- **Independent substrates** predicts no parameter transfer across kingdoms —
  contradicted by L3.

The framework is distinguished from competitors by the combination of T3 (biphasic
kinetics), T4–T5 (niche over Ne), T6 (integration-depth ordering), and
L3 (cross-kingdom transfer). No single test suffices; the framework's
strength is convergent evidence across multiple independent predictions.

# Function Map

```{r function-map, echo = FALSE, results = "asis"}
# Build the function map table
map <- data.frame(
  Function = c(
    "`pgls_orobanchaceae()`",
    "`pgls_cross_family()`",
    "`endosymbiont_biphasic()`",
    "`niche_vs_ne()`",
    "`pangenome_fluidity()`",
    "`gene_loss_ordering()`",
    "`ltee_cosegregation()`",
    "`transfer_test()`",
    "`threshold_model()`"
  ),
  `the framework's prediction` = c(
    "Plastome size ~ parasitism level",
    "Gradient replicates across families",
    "Biphasic (fast→slow) reduction",
    "Niche breadth > Ne",
    "Pan-genome ~ lifestyle, not Ne",
    "Integration-depth orders loss",
    "Loss co-segregates with sweeps",
    "Plant params predict bird ordering",
    "Biphasic: Phase 1 ≫ Phase 2 rate"
  ),
  `Competitor Prediction` = c(
    "Relaxed selection: same gradient",
    "Stochastic loss: same gradient",
    "Constant rate (Lynch); accelerating (Muller)",
    "Drift: Ne is primary driver",
    "Ne-only model",
    "Random loss: no ordering",
    "Independent assortment: 61.7%",
    "Substrates are independent",
    "Constant rate; accelerating"
  ),
  `Distinguishes the framework?` = c(
    "No",
    "No",
    "Yes",
    "Yes",
    "Yes",
    "Yes",
    "No",
    "Yes",
    "Yes"
  ),
  `Key Caveat` = c(
    "Also predicted by relaxed selection",
    "Also predicted by relaxed selection",
    "n = 10 genera; McCutcheon's model predicts same",
    "Tolerance data quality",
    "Ne data sparse in Dewar et al.",
    "Requires annotated gene categories",
    "Hitchhiking confound; reported as suggestive",
    "Ordering transfers; rate does not",
    "Model is deterministic, no stochastic component"
  ),
  check.names = FALSE
)

knitr::kable(map, caption = "Summary of the framework test functions, competitor predictions, and whether each test distinguishes the framework.")
```

Six of the seven empirical tests (T1–T7) support the framework's prediction. But
only **T3, T4, T5, and T6** distinguish the framework from their named competitors.
The cross-kingdom transfer test (L3) is the strongest: it distinguishes the framework
from all competitors and provides the most stringent falsification
criterion.

# Worked Example: T6 — Gene-Loss Ordering

This worked example walks through **T6: `gene_loss_ordering()`**, which
tests the framework's core prediction that integration depth determines the order in
which gene categories are lost during genome reduction.

## Step 1: Load the Required Packages

```{r t6-load}
library(valence.foundry)
```

## Step 2: Prepare the Data

The `gene_loss_ordering()` function requires a data frame with gene
categories, their integration-depth (dependency) scores, and the observed
loss ranks across one or more parasitic lineages.

The manuscript data for Orobanchaceae and Cuscuta (dodder) gene loss
produced the following dependency scores and loss ranks:

```{r t6-data}
gene_data <- data.frame(
  category = c("ndh", "rpo", "psa", "psb", "atp", "rpl_rps"),
  dependency_score = c(0, 1, 1, 2, 3, 5),
  orobanchaceae_loss_rank = c(1, 2, 2, 2, 3, 4),  # Orobanchaceae
  cuscuta_loss_rank = c(1, 2, 3, 4, 5, 6),          # Cuscuta
  stringsAsFactors = FALSE
)

str(gene_data)
```

The six categories represent functional gene groups:

| Category | Function | Dependency Score | Meaning |
|----------|----------|-----------------|---------|
| `ndh` | NADH dehydrogenase | 0 | Shallow — plastid-encoded, non-essential under heterotrophy |
| `rpo` | RNA polymerase | 1 | Substitutable — host provides |
| `psa` | Photosystem I | 1 | Substitutable |
| `psb` | Photosystem II | 2 | Moderately integrated |
| `atp` | ATP synthase | 3 | Deeply integrated — core metabolism |
| `rpl_rps` | Ribosomal proteins | 5 | Deepest — translation machinery |

Higher dependency scores reflect deeper functional integration. The framework predicts
that categories with higher scores are retained longer (higher loss rank
= lost later).

## Step 3: Run the Test

```{r t6-run, warning = FALSE, message = FALSE}
# The gene_loss_ordering function implements the exact permutation test
# described in the monograph (§12.4). With 6 categories, 6! = 720
# permutations exhaust all possible orderings.
result <- gene_loss_ordering(gene_data, seed = 42, n_perm = 720)
```

The `n_perm = 720` argument requests an **exact permutation test**: with 6
categories, 6! = 720 permutations exhaust all possible orderings. Setting
`seed = 42` ensures reproducibility (DFT Axiom A2).

## Step 4: Examine the Results

```{r t6-examine}
result$values
```

The output contains:

| Metric | Value | Interpretation |
|--------|-------|----------------|
| `spearman_rho` | ~0.96 | Very high positive correlation — deeply integrated functions are retained longer, exactly as the framework predicts |
| `permutation_p` | ~0.008 | Only about 6 of 720 possible permutations produce a correlation this extreme by chance |
| `pseudo_r_squared` | ~0.82 | Integration depth explains ~82% of the variance in loss ordering |
| `cross_family_concordance` | ~0.94 | Orobanchaceae and Cuscuta show nearly identical loss ordering — the pattern is not lineage-specific |

```{r t6-metadata}
result$metadata
```

## Step 5: Visualise the Ordering

```{r t6-plot, fig.width = 6, fig.height = 4, fig.cap = "Gene loss rank vs integration depth in Orobanchaceae. Each point is a gene category. The strong positive trend (ρ ≈ 0.96) supports the framework's integration-depth ordering prediction."}
plot(gene_data$dependency_score, gene_data$orobanchaceae_loss_rank,
     main = "Gene Loss Ordering: Orobanchaceae",
     xlab = "Integration Depth (Dependency Score)",
     ylab = "Loss Rank (1 = lost first, 6 = lost last)",
     pch = 19, cex = 1.5, col = "#1b9e77")
text(gene_data$dependency_score, gene_data$orobanchaceae_loss_rank,
     labels = gene_data$category, pos = 4, cex = 0.8)
abline(lm(orobanchaceae_loss_rank ~ dependency_score, data = gene_data),
       col = "#d95f02", lwd = 2)
```

## Step 6: Interpret

The results support the framework: integration depth predicts loss ordering with high
confidence (ρ ≈ 0.96, p ≈ 0.008). The competitor (random loss) predicts
ρ ≈ 0, which is firmly rejected. The cross-family concordance (~0.94)
shows the ordering is robust across independently evolved parasitic
lineages.

**What the test does NOT tell us:**
- It does not distinguish the framework from relaxed selection at the *rate* level
  (that is T3's job).
- It does not establish causation — only that integration depth is a
  strong predictor of retention order.
- The small number of categories (6) limits resolution; finer-grained
  analyses with more categories would strengthen the result.

# Limitations

## Statistical Limitations

1. **Small sample sizes**
   - T1 (Orobanchaceae): 12 species — borderline for PGLS; influential
     points can swing the result.
   - T3 (Endosymbionts): 10 genera — sensitivity to individual data points
     not fully tested; the biphasic fit is fragile.
   - T6 (Gene Loss): 6 categories — the permutation test is exact but
     coarse; the minimum achievable p-value is 1/720 ≈ 0.0014.

2. **Non-distinguishing tests**
   - **T1 (PGLS Orobanchaceae)** and **T2 (Cross-family replication)**
     predict the same gradient as relaxed selection (Lahti et al. 2009).
     These tests are necessary but not sufficient. They establish that the
     predicted pattern exists, but do not adjudicate between mechanisms.
   - **T7 (LTEE co-segregation)** is reported as suggestive only. The
     hitchhiking confound means that co-segregation with beneficial
     sweeps does not necessarily imply passive drift.

3. **Confounds**
   - **T7 hitchhiking**: loss mutations near beneficial sweeps could be
     dragged along by selection on linked sites, not by passive drift in
     unused genes. The arbitrary window (±2000 generations) and potential
     misspecification of the null (61.7% expected) further weaken the
     result.
   - **T3 vs McCutcheon**: metabolic complementarity (McCutcheon & Moran
     2012) predicts the same biphasic correlation as the framework. T3 alone cannot
     distinguish the two — that requires the cross-kingdom test (L3).

## Methodological Limitations

4. **Rate does not transfer across kingdoms**
   The cross-kingdom test (L3) shows that the *ordering* of loss is
   conserved across plants and birds, but the *rate* is not. The
   plant-derived slope predicts the bird ordering ($\rho = 0.755$,
   $p = 0.031$) but the actual rate of morphological change in birds
   differs from the rate of plastome reduction in plants. The framework claims
   universality of ordering, not of rate.

5. **Phylogenetic pseudoreplication in T3**
   The endosymbiont dataset includes multiple strains from the same genus
   (e.g., 25+ *Buchnera* strains). The function aggregates to genus means,
   reducing the effective sample size. A multilevel model would be more
   appropriate but is not implemented.

6. **No temporal data**
   All tests are cross-sectional. The framework dynamical model makes predictions
   about *rates* of change over time, but the empirical tests use
   contemporary measurements as proxies for evolutionary trajectories.
   T3's symbiosis age estimates are uncertain and may introduce error.

## What These Functions Cannot Establish

| Cannot establish | Why |
|-----------------|-----|
| That the framework is the *only* mechanism | Competitors overlap on some predictions |
| Causation | All tests are correlational (comparative method) |
| Irreversibility | The cusp catastrophe model is not yet empirically tested |
| Universal applicability | Only tested on plants, bacteria, and birds |
| Quantitative parameters | The formal model's parameters ($\lambda$, $\theta$, $\alpha$) are not yet fit to empirical data |

# Interpretation Guide

## Reading the Results

Every test function returns an **A6 proof object** — a list with two
components:

```{r interpret-structure, eval = FALSE}
result <- some_test_function(data)
names(result)
# [1] "values"   "metadata"
```

The `values` vector contains the statistical quantities (effect sizes,
p-values, fit metrics). The `metadata` list records the seed, sample size,
method, and convergence status.

## What Supports the framework

| Observation | Interpretation |
|-------------|----------------|
| Significant negative $\beta$ in T1 | Plastome shrinks with parasitism depth |
| High $|r|$ > 0.9 in T2 | Gradient is robust across families |
| Biphasic (logistic) model fits better than linear in T3 | Decelerating reduction consistent with threshold gating |
| Bayes factor > 3 in T3 | Positive evidence for logistic over exponential |
| niche_R² > ne_R² in T4–T5 | Niche breadth, not drift, drives gene loss |
| $\rho \gg 0$ and $p < 0.05$ in T6 | Integration depth orders loss |
| Less than 61.7% in T7 | Loss co-segregates less than expected by chance |
| $\rho > 0$ and $p < 0.05$ in L3 | Ordering principle transfers across kingdoms |

## What Refutes the framework

| Observation | Interpretation |
|-------------|----------------|
| Non-significant $\beta$ in T1 | No plastome–parasitism relationship |
| $|r| \approx 0$ in T2 | Gradient does not replicate |
| Linear model fits better than logistic in T3 | No biphasic kinetics |
| ne_R² ≥ niche_R² in T4–T5 | Drift dominates, not niche |
| $\rho \approx 0$ in T6 | No integration-depth ordering |
| $\rho \approx 0$ or negative in L3 | Cross-kingdom ordering fails |
| Observed ≥ expected rate in T7 | No co-segregation deficit |

## Handling INCONCLUSIVE Results

Several tests can produce ambiguous outcomes:

**T1/T2 (non-distinguishing).** Even if significant, these tests do not
distinguish the framework from relaxed selection. A significant result is *necessary*
for the framework but not *sufficient*. Report as "consistent with the framework and
with relaxed selection." The distinguishing weight must come from T3–T6.

**T3 (biphasic fit).** If the logistic model fails to converge, the
result is inconclusive — not a refutation. Try:
- Check the data quality (outliers, missing values)
- Increase `nls.control(maxiter = ...)` 
- Use a different start value for the logistic fit parameters
- Consider a spline-based approach outside the package

**T4/T5 (niche vs Ne).** If niche_R² ≈ ne_R², the result is ambiguous.
This could mean both mechanisms contribute or that the data lacks
resolution. Try:
- Check for collinearity between niche breadth and Ne
- Test with a reduced dataset excluding intermediate lifestyles
- Collect additional lifestyle-classified species

**T6 (permutation test).** With $n = 6$ categories, the permutation test
has limited resolution. If $p > 0.05$, the result is *inconclusive*, not
a refutation. The minimum achievable p-value is $1/720 \approx 0.0014$.
Consider:
- Adding more gene categories (finer functional annotation)
- Combining loss ranks across multiple lineages (increases effective N)
- Using the pseudo-R² as a continuous measure of support

**T7 (suggestive only).** The LTEE test is explicitly reported as
suggestive in the monograph (§12.5). A non-significant result does not
refute the framework. A *significant* result (observed ≥ expected) would be more
informative, as it would contradict the co-segregation deficit prediction.

**L3 (cross-kingdom).** With $n = 8$ bird structures, the Spearman test
has limited power. A non-significant result ($p > 0.05$) is inconclusive;
the expected $\rho$ under the alternative hypothesis is unknown. Report
the observed $\rho$ and its confidence interval.

## Summary Decision Tree

```
                     ┌─ T1–T2 significant?
                     │
                ┌────┴────┐
                │         │
               YES        NO ──→ the framework not supported 
                │              (but also doesn't refute)
           ┌────┴────┐
           │         │
      T3 logistic   T3 linear
      fits better   fits better ──→ Refutes biphasic prediction
           │
      ┌────┴────┐
      │         │
   T4–T5      T4–T5
 niche > Ne  Ne > niche ──→ Refutes niche-over-drift
      │
      │    ┌────┴────┐
      │    │         │
   T6 ρ ≫ 0 │     T6 ρ ≈ 0 ──→ Refutes integration-depth ordering
      │    │
      │    │    ┌────┴────┐
      │    │    │         │
      │    │ L3 ρ > 0  L3 ρ ≈ 0 ──→ Refutes cross-kingdom principle
      │    │    │
      │    │ ┌──┴──┐
      │    │ │    │
      │    │ the framework  Ambiguous ──→ More data needed
      │    │ supported
      │    │
      └────┴─────────────────── the framework strongly supported
```

In practice, the strength of the framework is that **multiple
independent tests converge on the same conclusion**. A single ambiguous
or even refuting result is informative but does not sink the framework
if the other tests are strong. The cross-kingdom test (L3) carries the
most weight because it is the most specific — no competitor predicts
parameter transfer across kingdoms.

## Reporting Convention

When reporting results, use the WCI (Weighted Confidence Index) scale
from the monograph:

| Tier | WCI Range | Meaning | Example |
|------|-----------|---------|---------|
| Tier 1 | > 80 | High confidence | T6 (ρ ≈ 0.96, exact permutation p = 0.008) |
| Tier 2 | 55–80 | Moderate confidence | L3 (ρ = 0.76, p = 0.031, n = 8) |
| Tier 3 | < 55 | Low confidence | T7 (suggestive, hitchhiking confound) |

Each test's WCI is computed from six dimensions: theoretical coherence,
empirical support, replicability, independent uptake, explanatory power,
and falsifiability. The vignette-level results in the monograph (§12)
provide the full WCI breakdown for each test.

# References

- Bobay, L.-M. & Ochman, H. (2017). *Impact of effective population size
  on genome evolution*. Nature Communications.
- Dewar, A. et al. (2024). *Pan-genome fluidity and lifestyle*. [Data collected
  for the framework monograph.]
- Good, B. H. et al. (2017). *The dynamics of molecular evolution over 60,000
  generations*. Nature, 551:45–50.
- Lahti, D. C. et al. (2009). *Relaxed selection in the wild*. Trends in
  Ecology & Evolution, 24:487–496.
- Leiby, N. & Marx, C. J. (2014). *Metabolic erosion primarily through
  mutation accumulation*. mBio, 5:e01967-14.
- Lynch, M. (2007). *The frailty of adaptive hypotheses for the origins of
  organismal complexity*. PNAS, 104:8597–8604.
- McCutcheon, J. P. & Moran, N. A. (2012). *Extreme genome reduction in
  symbiotic bacteria*. Nature Reviews Microbiology, 10:13–26.
- Ritch-Frel, J. & Phil, E. (2026). *the framework: A formal framework
  for capacity reallocation under niche-driven selection*. The framework Monograph.